Concurrent Programming with CXL

### 2020-06-28 book/paper.tex
Latest

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}
do not make a distinction between processes and threads.})
read and update shared variables
\index{shared variable}
concurrently and synchronize with one another makes programs more
complicated than programs where only one thing happens at a time.

@@

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

Model checking is not a replacement for formal verification.
\index{formal verification}
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be
simple or it will take longer than we may care to wait for or run
out of memory.
In particular, the model needs to have a relatively small number of
reachable states.

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}
operator provides the input to the program by selecting some value from the
set~$S$, while the $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

In CXL, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ cxl -c N=100 triangle.cxl
#states = 103
no issues found
$
\end{verbatim}
\end{code}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.  If not,
it reports the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}
in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

@@

The result in this particular interleaving is that \textit{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}
When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

\chapter{The CXL Virtual Machine}
\label{ch:cxlmachine}
\index{CXL Virtual Machine}

CXL programs are compiled to CXL bytecode
\index{bytecode}
(machine instructions for a virtual machine),
which in turn is executed by the CXL virtual machine (CVM).
\index{virtual machine}
\index{CXL Virtual Machine}
\index{CVM}
To understand the problem of concurrent computing, it
is important to have a basic understanding of the CVM.

@@

CXL programs, and indeed the CVM,  manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

Instead of bits and bytes, a CVM manipulates CXL values.
A CVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CVM instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CVM has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}
a program counter,
\index{program counter}
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}
that contains a CXL value.
\end{itemize}

A name tag consists of the name of the main method of the process,
along with an optional tag specified in the \texttt{spawn}
\index{spawn statement}
\index{tag}
statement.
The default tag is the first argument to the method.
A name tag is represented as a dictionary with keys \texttt{.name}
and \texttt{.tag}, but we shall often abbreviate it using the notation
\texttt{name/tag}.
For example, in Figure~\ref{fig:inc}, the created processes have name tags
\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}
\index{continuation}
it contains the value of
its name tag, program counter, stack, and register.
The state of a CVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% Figure~\ref{fig:incstate} shows an example of a reachable state for the program in Figure~\ref{fig:inc}.

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
\index{directory}
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}
of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
Each process in the CVM is predominantly a \emph{stack machine},
\index{stack machine}
but it also has a register.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process, with name tag \texttt{\_\_init\_\_/()},
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \textit{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}:
a set of instructions where only one process is allowed to execute at a
time.
\index{critical section}
\index{critical region}

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}
\index{thread}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes in the critical section.
This property is called \emph{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \textbf{atLabel}~$L$,
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \textbf{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \textbf{nametag}() operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

Mutual exclusion is an example of a \emph{safety property},
\index{safety property}
a property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}:
\index{liveness property}
we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if there exists a non-empty
set $S$ of processes that are trying to enter the critical section and
processes in the intersection always leave eventually, then eventually one
process in $S$ will enter the critical section.
We call this \emph{progress}.
\index{progress}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

Figure~\ref{fig:mutex} shows a high-level state diagram
\index{state diagram} of mutual exclusion and progress.
The circles represent states, while the arrows represent possible state
transitions (high-level steps).  The label in each circle summarizes the state;
the label on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate until after leaving the critical section.
We will now consider various approaches toward implementing this
specification.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Note that the word ``lock'' does not describe the concept well---it really
is more like a baton or token that only one process can hold.

@@

\chapter{Peterson's Algorithm}
\label{ch:peterson}
\index{Peterson's Algorithm}

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
at any point in time the CVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.

@@

The CVM starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

An \emph{invariant} of a program
\index{invariant}
is a predicate that holds over all states that are reachable by that program.
We want to show that exclusivity is an invariant.
In other words, we want to show that the set of reachable states of executing
the program
is a subset of the set of states where there is at most one process in the critical
section.

@@

To solve this problem, we will use what is called an
\emph{inductive invariant}.
\index{inductive invariant}
An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $\mathcal{I}$ holds in the initial state.
\item For any state in which $\mathcal{I}$ holds and any process in the
state takes a step, then $\mathcal{I}$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

@@

\chapter{CXL Methods and Pointers}
\label{ch:method}
\index{CXL method}

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If $x$ is a shared variable, then the expression \&$x$
 is a pointer to $x$ (also known as the \emph{address}
\index{address}
of $x$).
Conversely, if $p$ is a pointer to a shared variable, then the
expression \texttt{\^{}}$p$ is the value of the shared variable.
In particular, \texttt{\^{}\&}$x$ == $x$.
This is similar to how C pointers work.

@@

You can put the first three methods in its own CXL source file
and include it using the CXL \texttt{import} statement.
\index{import statement}
\index{module}
This would
make the code re-usable by other applications.

@@

\chapter{Spinlock}
\label{ch:spinlock}
\index{spinlock}

@@

\chapter{Locks and Blocking}
\label{ch:synch}
\index{lock}
\index{synch module}

@@

Here $L$ is a shared variable containing a list.
Another process can revive the process using the \textbf{go}
\index{go statement}
statement:

@@

\chapter{Semaphores}
\label{ch:semaphore}
\index{semaphore}

>>>>

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}%
do not make a distinction between processes and threads.})
read and update shared variables
\index{shared variable}%
concurrently and synchronize with one another makes programs more
complicated than programs where only one thing happens at a time.

@@

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}%
program is mostly \emph{deterministic}.
\index{determinism}%
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}%
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}%
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}%
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}%
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

Model checking is not a replacement for formal verification.
\index{formal verification}%
Formal verification proves that a program is correct.  Model checking only
verifies that a program is correct for some \emph{model}.  Think of
a model as a test program.  Because model checking tries every possible
execution, this means that the test program needs to be
simple or it will take longer than we may care to wait for or run
out of memory.
In particular, the model needs to have a relatively small number of
reachable states.

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}%
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}%
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}%
operator provides the input to the program by selecting some value from the
set~$S$, while the $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

In CXL, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ cxl -c N=100 triangle.cxl
#states = 103
no issues found
$
\end{verbatim}
\end{code}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.  If not,
it reports the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}%
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}%
in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

@@

The result in this particular interleaving is that \textit{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}%
When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

\chapter{The CXL Virtual Machine}
\label{ch:cxlmachine}
\index{CXL Virtual Machine}%

CXL programs are compiled to CXL bytecode
\index{bytecode}%
(machine instructions for a virtual machine),
which in turn is executed by the CXL virtual machine (CVM).
\index{virtual machine}%
\index{CXL Virtual Machine}%
\index{CVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of the CVM.

@@

CXL programs, and indeed the CVM,  manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}%
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}%
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

Instead of bits and bytes, a CVM manipulates CXL values.
A CVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CVM instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CVM has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}%
a program counter,
\index{program counter}%
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}%
that contains a CXL value.
\end{itemize}

A name tag consists of the name of the main method of the process,
along with an optional tag specified in the \texttt{spawn}
\index{spawn statement}%
\index{tag}%
statement.
The default tag is the first argument to the method.
A name tag is represented as a dictionary with keys \texttt{.name}
and \texttt{.tag}, but we shall often abbreviate it using the notation
\texttt{name/tag}.
For example, in Figure~\ref{fig:inc}, the created processes have name tags
\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a CVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% Figure~\ref{fig:incstate} shows an example of a reachable state for the program in Figure~\ref{fig:inc}.

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
\index{directory}%
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}%
of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
Each process in the CVM is predominantly a \emph{stack machine},
\index{stack machine}%
but it also has a register.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process, with name tag \texttt{\_\_init\_\_/()},
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \textit{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}:
a set of instructions where only one process is allowed to execute at a
time.
\index{critical section}%
\index{critical region}%

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}%
\index{thread}%

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}%
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes in the critical section.
This property is called \emph{mutual exclusion}.
\index{mutual exclusion}%
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \textbf{atLabel}~$L$,
\index{atLabel operator}%
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \textbf{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \textbf{nametag}() operator
\index{nametag operator}%
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

Mutual exclusion is an example of a \emph{safety property},
\index{safety property}%
a property that ensures that \emph{nothing bad will happen}, in this case
two processes being in the critical section.
What we need now a \emph{liveness property}:
\index{liveness property}%
we want to ensure that
\emph{eventually something good will happen}.
There are various possible liveness properties we could use,
but here we will propose the following informally: if there exists a non-empty
set $S$ of processes that are trying to enter the critical section and
processes in the intersection always leave eventually, then eventually one
process in $S$ will enter the critical section.
We call this \emph{progress}.
\index{progress}%

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

Figure~\ref{fig:mutex} shows a high-level state diagram
\index{state diagram} of mutual exclusion and progress.%
The circles represent states, while the arrows represent possible state
transitions (high-level steps).  The label in each circle summarizes the state;
the label on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate until after leaving the critical section.
We will now consider various approaches toward implementing this
specification.

@@

You may already have heard of the concept of a \emph{lock}
\index{lock}%
and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Note that the word ``lock'' does not describe the concept well---it really
is more like a baton or token that only one process can hold.

@@

\chapter{Peterson's Algorithm}
\label{ch:peterson}
\index{Peterson's Algorithm}%

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}%
at any point in time the CVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}%
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}%
Steps in CXL are atomic.

@@

The CVM starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}%
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}%
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}%
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

An \emph{invariant} of a program
\index{invariant}%
is a predicate that holds over all states that are reachable by that program.
We want to show that exclusivity is an invariant.
In other words, we want to show that the set of reachable states of executing
the program
is a subset of the set of states where there is at most one process in the critical
section.

@@

To solve this problem, we will use what is called an
\emph{inductive invariant}.
\index{inductive invariant}%
An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $\mathcal{I}$ holds in the initial state.
\item For any state in which $\mathcal{I}$ holds and any process in the
state takes a step, then $\mathcal{I}$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

@@

\chapter{CXL Methods and Pointers}
\label{ch:method}
\index{CXL method}%

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}%
to shared variables.
If $x$ is a shared variable, then the expression \&$x$
 is a pointer to $x$ (also known as the \emph{address}
\index{address}%
of $x$).
Conversely, if $p$ is a pointer to a shared variable, then the
expression \texttt{\^{}}$p$ is the value of the shared variable.
In particular, \texttt{\^{}\&}$x$ == $x$.
This is similar to how C pointers work.

@@

You can put the first three methods in its own CXL source file
and include it using the CXL \texttt{import} statement.
\index{import statement}%
\index{module}%
This would
make the code re-usable by other applications.

@@

\chapter{Spinlock}
\label{ch:spinlock}
\index{spinlock}%

@@

\chapter{Locks and Blocking}
\label{ch:synch}
\index{lock}%
\index{synch module}%

@@

Here $L$ is a shared variable containing a list.
Another process can revive the process using the \textbf{go}
\index{go statement}%
statement:

@@

\chapter{Semaphores}
\label{ch:semaphore}
\index{semaphore}%
